---
title: "Chicago analysis"
output: html_document

format:
  html:
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

library(tidyverse)
library(readr)
library(ggplot2)

chicago_data <- read_csv("./chicago_listings.csv")
head(chicago_data)
print(colnames(chicago_data))
```
```{r}
#View the summary of the tibble
str(chicago_data)
```
```{r}
clean_data <- function(df){
  
  # Select desired columns
  new_df <- df |> select(id, host_id, host_response_time, host_response_rate, host_is_superhost, neighbourhood_cleansed, property_type, room_type, accommodates, bathrooms, bedrooms, beds, amenities, price, minimum_nights, availability_30, availability_365, number_of_reviews, last_review, review_scores_rating, review_scores_location, reviews_per_month, longitude, latitude, host_identity_verified, host_total_listings_count)
  # Convert price to numeric after removing the dollar sign, and commas
  new_df$price <- as.numeric(gsub("[$,]", "", new_df$price))

  # Convert host_response_rate to numeric after removing the '%' sign
  new_df$host_response_rate <- as.numeric(gsub("%", "", new_df$host_response_rate))
  
  # Drop rows where price is null
  new_df <- new_df |> filter(!is.na(price))


  # One-hot encode 'amenities' column
  # new_df <- new_df %>%
  # mutate(amenities_list = strsplit(amenities, ",\\s*")) %>% # Split by comma and optional space
  # unnest_wider(amenities_list, names_sep = "_") %>%        # Spread into multiple columns
  # pivot_longer(cols = starts_with("amenities_list"),       # Reshape to long format for processing
  #              names_to = NULL, values_to = "amenity") %>%
  # filter(!is.na(amenity)) %>%                              # Remove NA values
  # mutate(amenity = trimws(amenity)) %>%                   # Remove leading/trailing spaces
  # pivot_wider(names_from = amenity,                       # One-hot encode
  #             values_from = amenity,
  #             values_fill = 0,
  #             values_fn = function(x) 1) %>%
  # group_by(id) %>%                                        # Reaggregate by ID
  # summarise(across(everything(), max, na.rm = TRUE))
  
  return(new_df)
}

new_chicago <- clean_data(chicago_data)
print(head(new_chicago))

#View percentage of missing values per column
colSums(is.na(new_chicago)) / nrow(new_chicago)
unique_neighbourhoods <- chicago_data %>%
  distinct(neighbourhood_cleansed) %>%
  pull(neighbourhood_cleansed)
print(unique_neighbourhoods)

```

**EDA**
```{r}
# Price density with and without outliers (2 plots total)
# With outliers
ggplot(new_chicago, aes(x = price)) +
  geom_density(color="forestgreen") +
  labs(x = "Price", y = "Density", title = "Price Density with Outliers")

# Without outliers using the IQR method
q1 <- quantile(new_chicago$price, 0.25)
q3 <- quantile(new_chicago$price, 0.75)
IQR <- q3 - q1
# Define outlier bounds
lower_bound <- q1 - 1.5 * IQR
upper_bound <- q3 + 1.5 * IQR

new_chicago_wo_price_outliers <- new_chicago %>% filter(price >= lower_bound & price <= upper_bound)

# Combine datasets for plotting
new_chicago$outlier_status <- "With Outliers"
new_chicago_wo_price_outliers$outlier_status <- "Without Outliers"
combined_data <- rbind(new_chicago, new_chicago_wo_price_outliers)

custom_colors <- c("With Outliers" = "#FF6F61", "Without Outliers" = "#6CA0DC")

# Plot density for both datasets
ggplot(combined_data, aes(x = price, fill = outlier_status)) +
  geom_density(alpha = 0.7) +
  facet_wrap(~outlier_status, scales = "free") +
  scale_x_continuous(labels = scales::dollar_format()) + # Format x-axis as currency
  scale_fill_manual(values = custom_colors) + # Use custom colors
  labs(
    title = "Chicago Rental Price Density: With and Without Outliers",
    x = "Price (USD)",
    y = "Density",
    fill = "Data Type"
  ) +
  theme_minimal(base_family = "Arial", base_size = 14) + # Cleaner font and base size
  theme(
    panel.grid = element_blank(), # Remove grid lines
    axis.line = element_line(color = "black"), # Add axis lines
    strip.text = element_text(size = 14, face = "bold"), # Style facet labels
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), # Title styling
    legend.position = "none"  # Remove the legend
  )

```
```{r}
print(mean(new_chicago$price))
print(mean(new_chicago_wo_price_outliers$price))
```


```{r}
# Average $ by neighborhood
avg_price_by_neighborhood <- new_chicago %>% 
  group_by(neighbourhood_cleansed) %>% 
  summarise(average_price = mean(price, na.rm = TRUE)) %>% 
  arrange(desc(average_price))

print(avg_price_by_neighborhood)

max_price_neighborhood <- avg_price_by_neighborhood %>% 
  filter(average_price == max(average_price)) %>%
  select(neighbourhood_cleansed, average_price)

min_price_neighborhood <- avg_price_by_neighborhood %>% 
  filter(average_price == min(average_price)) %>%
  select(neighbourhood_cleansed, average_price)

# Print the results
print(max_price_neighborhood)
print(min_price_neighborhood)

top_10_expensive_neighborhood <- head(avg_price_by_neighborhood, 10)

ggplot(top_10_expensive_neighborhood, aes(x = reorder(neighbourhood_cleansed, average_price), y = average_price)) +
  geom_bar(stat = "identity", fill = "lightblue", color = "black", alpha = 0.7) +
  ggtitle("Chicago: Top 10 Neighborhoods with Highest Mean Prices") +
  xlab("Neighborhood") +
  ylab("Average Price") +
  coord_flip() +
  theme_minimal(base_family = "Arial", base_size = 12)
```

```{r}
# Average price by property type
avg_price_by_property_type <- new_chicago |>
  group_by(property_type) |>
  summarise(average_price = mean(price, na.rm = TRUE))

print(avg_price_by_property_type)

# Price per property type
# Grouping into broader category groups
new_chicago_wo_price_outliers <- new_chicago_wo_price_outliers %>%
  mutate(
    property_group = case_when(
      grepl("condo|loft|serviced apartment|aparthotel|rental unit|guest suit", property_type, ignore.case = TRUE) ~ "Apartment",
      grepl("home|bungalow|guesthouse|townhouse|vacation home|tiny home|cottage|villa", property_type, ignore.case = TRUE) ~ "House",
      grepl("private room", property_type, ignore.case = TRUE) ~ "Private Room",
      grepl("shared room", property_type, ignore.case = TRUE) ~ "Shared Room",
      TRUE ~ "Other"
    )
  )

# Count each group to verify
table(new_chicago_wo_price_outliers$property_group)

# Boxplot for aggregated property groups with custom axis splines
ggplot(new_chicago_wo_price_outliers, aes(x = property_group, y = price, fill = property_group)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 21, outlier.size = 2) +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c(
    "Apartment" = "#6CA0DC",
    "House" = "#FF6F61",
    "Private Room" = "#FFD700",
    "Shared Room" = "#28B463",
    "Other" = "#A569BD"
  )) +
  labs(
    title = "Chicago: Boxplot of Airbnb Prices by Property Types",
    x = "Property Type",
    y = "Price (USD)"
  ) +
  theme_minimal(base_family = "Arial", base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "none",
    axis.line = element_blank(), # Remove all default axis lines
    axis.line.x.bottom = element_line(color = "black", size = 0.8), # Add black line to bottom x-axis
    axis.line.y.left = element_line(color = "black", size = 0.8) # Add black line to left y-axis
  )
```

```{r}
# Boxplot of price by room type

# Custom color palette for room types
custom_colors <- c("Entire home/apt" = "#FF6F61", 
                   "Private room" = "#6CA0DC", 
                   "Shared room" = "#FFD700", 
                   "Hotel room" = "#7DCEA0")

ggplot(new_chicago_wo_price_outliers, aes(x = room_type, y = price, fill = room_type)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 21, outlier.size = 2) +
  scale_y_continuous(labels = scales::dollar_format()) + # Format y-axis as currency
  scale_fill_manual(values = custom_colors) + # Use custom colors for room types
  labs(
    title = "Chicago: Boxplot of Airbnb Prices by Room Type",
    x = "Room Type",
    y = "Price (USD)",
    fill = "Room Type"
  ) +
  theme_minimal(base_family = "Arial", base_size = 14) + # Cleaner font and base size
  theme(
    panel.grid = element_blank(), # Remove grid lines
    axis.line = element_line(color = "black"), # Add axis lines
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), # Title styling
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"), # Rotate x-axis labels
    legend.position = "none" # Remove legend (optional for a cleaner look)
  )

```
```{r}
ggplot(data = new_chicago_wo_price_outliers, aes(x = review_scores_rating, y = price)) +
  geom_point() +
  labs(title = "Price v. Rating with outliers",
       x = "Review Rating",
       y = "Price")


new_chicago_ratings_grouped <- new_chicago_wo_price_outliers %>%
  mutate(rating_category = case_when(
    review_scores_rating >= 0 & review_scores_rating < 2 ~ "Low (0-2)",
    review_scores_rating >= 2 & review_scores_rating < 4 ~ "Medium (2-4)",
    review_scores_rating >= 4 & review_scores_rating <= 5 ~ "High (4-5)"
  ))  


new_chicago_ratings_grouped <- new_chicago_ratings_grouped %>%
  filter(!is.na(rating_category))

ggplot(new_chicago_ratings_grouped, aes(x = rating_category, y = price)) +
  geom_boxplot(fill = "maroon", alpha = 0.7) +
  ggtitle("Boxplot of Price by Rating Groups") +
  xlab("Rating Category") +
  ylab("Price") +
  theme_minimal()
```

```{r}
new_chicago_expanded <- new_chicago_wo_price_outliers %>%
  filter(if_all(c(accommodates, bathrooms, bedrooms, beds), ~ !is.na(.))) %>% # Drop rows with NAs in specified columns by applying is.na
  mutate(
    accommodates = cut(accommodates, breaks = c(0, 2, 4, 6, 8, 10, Inf), labels = c("1-2", "3-4", "5-6", "7-8", "9-10", "11+")),
    bathrooms = cut(bathrooms, breaks = c(0, 1, 2, 3, 4, Inf), labels = c("0", "1", "2", "3", "4+")),
    bedrooms = cut(bedrooms, breaks = c(0, 1, 2, 3, 4, Inf), labels = c("0", "1", "2", "3", "4+")),
    beds = cut(beds, breaks = c(0, 1, 2, 3, 4, Inf), labels = c("0", "1", "2", "3", "4+"))
  ) %>%
  select(price, accommodates, bathrooms, bedrooms, beds) %>%
  pivot_longer(cols = c(accommodates, bathrooms, bedrooms, beds),
               names_to = "feature",
               values_to = "value")


# Create the faceted boxplot
ggplot(new_chicago_expanded, aes(x = factor(value), y = price)) +
  geom_boxplot(aes(fill = feature), color = "black", outlier.shape = NA) +
  facet_wrap(~ feature, scales = "free_x") +
  labs(title = "Chicago: Price by Property Features",
       x = "Feature Value",
       y = "Price (USD)") +
  theme_minimal() +
  theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none",
      panel.grid = element_blank(), # Remove grid lines
      axis.line = element_line(color = "black")
      ) +
  scale_fill_manual(values = c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3"))
```

```{r}
# Boxplot per Superhost Status
new_chicago_wo_superhost_nulls <- new_chicago_wo_price_outliers %>%
  filter(!is.na(host_is_superhost))

median_price_superhost <- new_chicago_wo_superhost_nulls %>%
  filter(host_is_superhost == TRUE) %>%
  summarize(median_price = median(price))

print(median_price_superhost)

median_price_nonsuperhost <- new_chicago_wo_superhost_nulls %>%
  filter(host_is_superhost == FALSE) %>%
  summarize(median_price = median(price))

print(median_price_nonsuperhost)

ggplot(new_chicago_wo_superhost_nulls, aes(x = host_is_superhost, y = price)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  labs(title = "Chicago: Price Differences by Superhost Status",
       x = "Superhost Status",
       y = "Price (USD)") +
  theme_minimal()
head(new_chicago_wo_superhost_nulls)
```

## Statistical Testing ##

**ANOVA of price by neighborhood**

**Null Hypothesis** $(H_0)$: No significant difference between the mean listing price across Chicago neighborhoods on Airbnb.

**Alternative Hypothesis** $(H_1)$: There is at least one neighborhood mean listing price that is different from the other Chicago neighborhood.


```{r}
aov_price <- aov(price ~ neighbourhood_cleansed, data = new_chicago_wo_price_outliers)
summary(aov_price)

```
- Conclusion: As the p-value < 0.05, we can reject the null hypothesis and conclude that there is a significant difference between at least two groups in the mean price listings across Chicago neighborhoods. 

**Price v. Property Type**

```{r}
aov_price_property <- aov(price ~ property_group, data = new_chicago_wo_price_outliers)
summary(aov_price_property)

```
- Conclusion: As the p-value < 0.05, we can reject the null hypothesis and conclude that there is a significant difference in mean listing prices between at least two property types. 

**2. T-test difference in mean price by superhost** 

**Null Hypothesis** $(H_0)$: No difference in the mean listing price between non-superhosts and superhosts on Airbnb.
$H_0: \mu_{\text{Non-Superhost}} = \mu_{\text{Superhost}}$

**Alternative Hypothesis** $(H_1)$: The mean listing prices by non-superhosts are less than those of superhosts on Airbnb.
$H_1: \mu_{\text{Non-Superhost}} \lt \mu_{\text{Superhost}}$

```{r}
t_test_result <- t.test(
  price ~ host_is_superhost,
  data = new_chicago_wo_superhost_nulls,
  alt = "less"
)
print(t_test_result)
```

- Conclusion: As the p-value < 0.05, we can reject the null hypothesis. There is enough evidence to conclude that the mean listing prices by non-superhosts are less than those of superhosts on Airbnb. We are 95% confident that the mean listing price for non-superhosts are on average at least $7.68 less than the mean listing price for superhosts on Airbnb. 


**3. Correlation test for mean price by response rate** 

```{r}
#Correlation test in mean price by response rate


listing_clean1 <- new_chicago_wo_price_outliers %>%
  mutate(host_response_rate = host_response_rate / 100)  

corr_price_response_rate <- cor.test(listing_clean1$price, listing_clean1$host_response_rate, method = "pearson")

print(corr_price_response_rate)
```
- Conclusion: Since the p-value < 0.05, we can reject the null hypothesis and there is enough evidence to conclude that the true correlation between the mean listing price and the host response rate is not equal to 0. The calculated correlation coefficient is 0.1054 which describes a weak, positive relationship. We are 95% confident that the true Pearson correlation coefficient between mean listing price and host response rate between 0.0819 and 0.1288. This ultimately means that there is not a strong relationship between a listing's price and the host's response rate. 

**4. Correlation test for Price vs. Review Score** 

```{r}
listing_clean2 <- new_chicago_wo_price_outliers %>%
  filter(!is.na(review_scores_rating))

corr_price_review_rating <- cor.test(listing_clean2$price, listing_clean2$review_scores_rating, method = "pearson")

print(corr_price_review_rating)

```
- Conclusion: Since the p-value < 0.05, we can reject the null hypothesis and there is enough evidence to conclude that the true correlation between the mean listing price and the listing's review rating is not equal to 0. The calculated correlation coefficient is 0.0801 which describes a weak, positive relationship. We are 95% confident that the true Pearson correlation coefficient between mean listing price and host response rate between 0.0546 and 0.1055. This ultimately means that there is not a strong relationship between a listing's price and the listing's review rating.

**5. Correlation test for Price vs. Accommodates** 

```{r}
listing_clean3 <- new_chicago_wo_price_outliers %>%
  filter(!is.na(accommodates))

corr_price_accommodates <- cor.test(listing_clean3$price, listing_clean3$accommodates, method = "pearson")

print(corr_price_accommodates)

```
- Conclusion: Since the p-value < 0.05, we can reject the null hypothesis and there is enough evidence to conclude that the true correlation between the mean listing price and the listing's guest capacity is not equal to 0. The calculated correlation coefficient is 0.491 which describes a moderate, positive relationship. We are 95% confident that the true Pearson correlation coefficient between mean listing price and host response rate between 0.473 and 0.508. Therefore, we can expect price to increase as the max. number of guests the listing can accommodate increases.

**6. Correlation test for Price vs. Bathrooms** 

```{r}
listing_clean4 <- new_chicago_wo_price_outliers %>%
  filter(!is.na(bathrooms))

corr_price_bathrooms <- cor.test(listing_clean4$price, listing_clean4$bathrooms, method = "pearson")

print(corr_price_bathrooms)
```
- Conclusion: Since the p-value < 0.05, we can reject the null hypothesis and there is enough evidence to conclude that the true correlation between the mean listing price and the listing's total bathrooms is not equal to 0. The calculated correlation coefficient is 0.334 which describes a weak, positive relationship. We are 95% confident that the true Pearson correlation coefficient between mean listing price and host response rate between 0.313 and 0.354. 

**7. Chi-squared association between Superhost Status and Room Type**

**Null Hypothesis** $(H_0)$: There is no association between Superhost status and the room type of Airbnb listings.

**Alternative Hypothesis** $(H_1)$: There is an association between Superhost status and the room type of Airbnb listings.


```{r}
listing_clean5 <- new_chicago_wo_price_outliers %>%
  filter(!is.na(host_is_superhost)) %>%
  filter(!is.na(room_type))

superhost_room_table <- table(listing_clean5$host_is_superhost, listing_clean5$room_type)

chi_squared_result <- chisq.test(superhost_room_table)
print(superhost_room_table)
print(chi_squared_result)
```
- Conclusion: As the p-value < 0.05, we can reject the null hypothesis and conclude that there is an association between Superhost status and the room type of the Airbnb listing.

**8. Price. Distance to City Center: ANOVA and t-test**

- ANOVA

**Null Hypothesis** $(H_0)$: There is no significant difference in mean Airbnb listing prices across distance to city center (Near, Medium, Far)

**Alternative Hypothesis** $(H_1)$: There is at least one distance category (Near, Medium, Far) in which the mean Airbnb listing price significantly differs from the rest

Conclusion: As the p-value < 0.05, we reject the null hypothesis as there is enough evidence to conclude that there is at least one distance category which the mean price differs significantly.

- t-test

**Null Hypothesis** $(H_0)$: No difference in the mean listing price of near and far Airbnb listings.
$H_0: \mu_{\text{Near}} = \mu_{\text{Far}}$

**Alternative Hypothesis** $(H_1)$: The mean listing prices of listings closer to the city center are greater than those of listings further away.
$H_1: \mu_{\text{Near}} \gt \mu_{\text{Far}}$
Conclusion: As the p-value > 0.05, we fail to reject the null hypothesis. As the 95% confidence interval does include 0, the difference between the mean listing prices of Near and Far listings are not statistically significant. 

```{r}
# Eucledian distance from city center
#Coordinates of the Loop in Chicago
ref_lat = 41.8820
ref_lon = -87.6278

euclidean_distance <- function(lat1, lon1, lat2, lon2) {
  sqrt((lat2 - lat1)^2 + (lon2 - lon1)^2)
}

#calculate distances from Chicago center
new_chicago_wo_price_outliers$distance_to_city_center <- mapply(
  euclidean_distance,
  lat1 = ref_lat, lon1 = ref_lon,
  lat2 = new_chicago_wo_price_outliers$latitude, lon2 = new_chicago_wo_price_outliers$longitude
)

# Create distance bins
new_chicago_wo_price_outliers$distance_bin <- cut(
  new_chicago_wo_price_outliers$distance_to_city_center,
  breaks = 3,
  labels = c("Near", "Medium", "Far")
)

# ANOVA
anova_price_v_distance <- aov(price ~ distance_bin, data = new_chicago_wo_price_outliers)
summary(anova_price_v_distance)
# Run Tukey to see intra bucket mean differences
TukeyHSD(anova_price_v_distance)

# separate into near and far and do t-test
new_chicago_wo_price_outliers$distance_category <- ifelse(new_chicago_wo_price_outliers$distance_to_city_center <= median(new_chicago_wo_price_outliers$distance_to_city_center), "Near", "Far")

t.test(price ~ distance_category, data = new_chicago_wo_price_outliers, alt = "greater")
```

## Crime
```{r}
chicago_crime_data <- read_csv("./chicago_crime2024.csv")
head(chicago_crime_data)
# Calculate total data records by District
district_crime_counts <- chicago_crime_data %>%
  group_by(District) %>%
  summarize(Total_Records = n())

# Print the result
print(district_crime_counts)
```

```{r}
# Get unique Description values
unique_descriptions <- chicago_crime_data %>%
  distinct(Description)

# Print the unique descriptions
print(unique_descriptions)

```

